Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2011-02-10 intro.tex
Non-dimensionalized equations, some minor fixes

In this study we will model IPMC materials via a multiphysics coupled problem 
consisting of the Poisson and Nernst-Planck equations (abbreviated by PNP in
the following). These equations are used to model charge transport in materials 
that includes ionic migration, diffusion, and convection. The charge transport 
process is a key mechanism for electromechanical transduction.

The PNP system is highly nonlinear and for a typical domain with two
electrodes, largest differences in charge concentration occur in a very narrow
region near the boundary. The computing power required for a full scale problem 
is significant. This is why we are interested in exploring adaptive algorithms
-- we hope to obtain meshes that are optimal in terms of calculation time and 
calculation error.

>>>>

In this study we will model IPMC materials via a multiphysics coupled problem 
consisting of the Poisson and Nernst-Planck equations (abbreviated by PNP in
the following). These equations are used to model charge transport in materials 
that include ionic migration, diffusion, and convection. The charge transport 
process is a key mechanism for electromechanical transduction.

The PNP system is highly nonlinear and for a typical domain with two
electrodes, largest differences in charge concentration occur in a very narrow
region near the boundary. The computing power required for a full scale problem 
is significant~\cite{pugal2010spie3d}. This is why we are interested in exploring adaptive algorithms
-- we hope to obtain meshes that are optimal in terms of calculation time and 
calculation error.

### 2011-02-10 model.tex
Non-dimensionalized equations, some minor fixes

  X=\frac xL,\ \ \
  Y=\frac yL,\ \ \
  \tau=\frac{tD}{\lambda_D L},\ \ \

@@

  \frac{DC_0}{\lambda_D L}\frac{\partial c}{\partial \tau} + \frac 1L \nabla_d\cdot
    \left(-\frac{DC_0}{L}\nabla_dc-c\frac{DC_0}{L}\nabla_d\varphi\right) &=& 0\label{eq:dim1}\\
  - \frac{\varepsilon R T}{L^2F^2C_0}\nabla_d\varphi&=&c-1\label{eq:dim2},

@@

  \epsilon=\frac{\lambda_D}{L},

@@

  \frac{\partial c}{\partial \tau} + \epsilon \nabla_d\cdot
    \left(-\nabla_d c - c \nabla_d \varphi\right) &=&0,\label{eq:dimensionless-nernst-planck}\\
  -2\epsilon^2\nabla_d\varphi&=&c-1\label{eq:dimensionless-poisson}.

@@

$$H^1\left(\Omega\right)=\left\{v\in L^2\left(\Omega\right);\ \nabla v \in \left[L^2\left(\Omega\right)\right]^2\right\}.

@@

  \int_{\Omega}\frac{\partial C}{\partial t}v^C d\mathbf{x}
  -\int_{\Omega}D\nabla^2Cv^C d\mathbf{x}-\int_{\Omega}K\nabla C\cdot\nabla\phi v^C d\mathbf{x}
   - \int_{\Omega}KC\nabla^2\phi v^C d\mathbf{x}=0.

@@

 \int_{\Omega}\frac{\partial C}{\partial t}v^C d\mathbf{x}+
  D\int_{\Omega}\nabla C\cdot\nabla v^C d\mathbf{x}-
  K\int_{\Omega}\nabla C \cdot \nabla \phi v^C d\mathbf{x}

@@

  + K\int_{\Omega}\nabla\left(Cv^C\right)\cdot \nabla \phi d\mathbf{x}
  -D\int_{\partial\Omega}\frac{\partial C}{\partial n}v^C d\mathbf{S}-
  \int_{\partial\Omega}K\frac{\partial\phi}{\partial n}Cv^C d\mathbf{S}=0.

@@

\eqref{eq:nernst-planck-boundary-2}, we have

@@

  \int_{\Omega}\frac{\partial C}{\partial t}v^C d\mathbf{x}+
  D\int_{\Omega}\nabla C \cdot \nabla v^C d\mathbf{x}-
  K\int_{\Omega}\nabla C \cdot \nabla \phi v^C d\mathbf{x}

@@

  + K\int_{\Omega}\nabla \phi \cdot \nabla C v^C d\mathbf{x}+
  K\int_{\Omega} C \left(\nabla\phi\cdot\nabla v^C\right) d\mathbf{x}=0.

@@

After the second and third terms cancel out, we obtain the final weak form of 

@@

  \int_{\Omega}\frac{\partial C}{\partial t}v^C d\mathbf{x}+
  D\int_{\Omega}\nabla C \cdot \nabla v^C d\mathbf{x}+
  K\int_{\Omega} C \left(\nabla\phi\cdot\nabla v^C\right) d\mathbf{x}=0.

@@

Analogously we derive also the weak form of the Poisson equation (\ref{eq:poisson-2}),

@@

  -\int_{\Omega}\nabla^2\phi v^\phi d\mathbf{x}-\int_{\Omega}LCv^\phi d\mathbf{x}+
  \int_{\Omega}LC_{0}v^\phi d\mathbf{x}=0.

@@

  \int_{\Omega}\nabla\phi\cdot\nabla v^\phi d\mathbf{x}-\int_{\Omega}LCv^\phi d\mathbf{x}+
  \int_{\Omega}LC_{0}v^\phi d\mathbf{x}=0.

@@

\subsection{Jacobian matrix and residual vector for the Newton's method}
To employ the Newton's method for the nonlinear system \eqref{eq:nernst-planck-weak-final},
\eqref{eq:poisson-weak-final}, formulas for the Jacobian matrix and residual vector need to
be derived. Time discretization will be performed using the second-order Crank-Nicolson 
method. The unknown solution components 
$C^{n+1}$ and $\phi^{n+1}$ at the end of the time step are expressed 
as linear combinations of finite element basis functions $v_k^{C}$ 
and $v_k^{\phi}$ with unknown coefficients,
\begin{eqnarray}
  C^{n+1} = C(Y^{n+1}) &=& \sum_{k=1}^{N^C} y_k^{C} v_k^{C}, \label{eq:cnotation}\\
  \phi^{n+1} = \phi(Y^{n+1}) &=& \sum_{k=1}^{N^{\phi}} y_k^{\phi} v_k^{\phi}\label{eq:phinotation}.
\end{eqnarray}
Here $Y^{n+1}$ is a coefficient vector of length $N^C + N^{\phi}$ comprising the unknown 
solution coefficients $y_k^{C}$ and $y_k^{\phi}$ (in this order). We will also be 
using $C^n = C(Y^n)$ and $\phi^n = \phi(Y^n)$ for the previous time step solutions.

With the notation~\eqref{eq:cnotation}, \eqref{eq:phinotation}, the time discretized 
Eq.~\eqref{eq:nernst-planck-weak-final} leads to the formula for the 
first part $F^C$ of the residual vector $F$,
\begin{eqnarray}
  F_i^C\left(Y\right) & = & \int_{\Omega} \frac{C(Y)}{\tau}v_i^C d\mathbf{x} - 
  \int_{\Omega} \frac{C^{n}}{\tau}v_i^C d\mathbf{x}\nonumber\\
  &&+\frac 12 \left[D\int_{\Omega} \nabla C(Y) \cdot \nabla v_i^C d\mathbf{x}+ 
  	D\int_{\Omega} \nabla C^{n} \cdot \nabla v_i^C d\mathbf{x}\right]\nonumber\\
  &&+ \frac 12 \left[K\int_{\Omega}C(Y) \left(\nabla \phi(Y) \cdot \nabla v_i^C\right) d\mathbf{x}+
  K\int_{\Omega}C^{n} \left(\nabla \phi^{n} \cdot \nabla v_i^C\right) d\mathbf{x}\right]\label{eq:Fc}
\end{eqnarray}
where $i = 1, 2, \ldots, N^C$.
Analogously, Eq.~\eqref{eq:poisson-weak-final} defines the second part $F^{\phi}$ 
of the residual vector $F$,
\begin{equation}
  F_i^{\phi}\left(Y\right) = \int_{\Omega} \nabla \phi(Y) \cdot \nabla v_i^{\phi} d\mathbf{x} 
  - \int_{\Omega} LC(Y)v_i^{\phi} d\mathbf{x} + \int_{\Omega} LC_0 v_i^{\phi} d\mathbf{x}
  \label{eq:Fphi}
\end{equation}
where $i = N^C + 1, N^C + 2, \ldots, N^C + N^{\phi}$. The nonlinear discrete problem 
that needs to be solved at the end of each time step thus has the form 
$F(Y) = 0$.

\newpage

\noindent
The Jacobian matrix $J(Y) = DF/DY$ has a $2\times 2$ block structure,
\begin{equation}
J(Y) = \left(
\begin{array}{cc}
  \displaystyle \frac{\partial F_i^C}{\partial y_j^C} & \displaystyle \frac{\partial F_i^C}{\partial y_j^{\phi}} \\ 
  \displaystyle \frac{\partial F_i^{\phi}}{\partial y_j^C} & \displaystyle \frac{\partial F_i^{\phi}}{\partial y_j^{\phi}}
\end{array}
\right),
\end{equation}
and its entries are obtained by calculating the partial derivatives of $F$ with
respect to the components of the coefficient vector $Y$. For this it is useful to 
realize that 
$$
\frac{\partial C(Y)}{\partial y_j^C} = v_j^C, \ \ \ 
\frac{\partial \nabla C(Y)}{\partial y_j^C} = \nabla v_j^C,\ \ \ \mbox{etc.}.
$$
We obtain
\begin{eqnarray}
  \frac{\partial F_i^C}{\partial y_j^C}(Y) &=& 
  \int_{\Omega} \frac{1}{\tau} v_j^C v_i^C d\mathbf{x} + 
  \frac 12 D\int_{\Omega} \nabla v_j^C \cdot \nabla v_i^C d\mathbf{x}
  + \frac 12 K\int_{\Omega} v_j^C \left(\nabla \phi(Y) \cdot \nabla v_i^C\right) d\mathbf{x},\label{eq:dFcdyc}\\
  \frac{\partial F_i^C}{\partial y_j^{\phi}}(Y) &=&
  \frac 12 K \int_{\Omega} C(Y) \left(\nabla v_j^{\phi} \cdot \nabla v_i^C\right) d\mathbf{x},\\
  \frac{\partial F_i^{\phi}}{\partial y_j^C}(Y) &=&
  - \int_{\Omega} L v_j^C v_i^{\phi} d\mathbf{x},\\
  \frac{\partial F_i^{\phi}}{\partial y_j^{\phi}}(Y) &=&
  \int_{\Omega} \nabla v_j^{\phi} \cdot \nabla v_i^{\phi} d\mathbf{x}\label{eq:dFphidyphi}.
\end{eqnarray}

\subsection{Newton's iteration}
At the beginning of the $(n+1)$st time step we 
set $Y^{n+1}_0 = Y^n$, where $Y^n$ is the coefficient vector  
that was calculated in the $n$th time step (or coming from the 
initial condition if $n = 0$). We set $k = 0$ and run the 
Newton's iteration 
\begin{eqnarray*}
J(Y^{n+1}_k) \delta Y^{n+1}_{k+1} &=& -F(Y^{n+1}_k), \\ 
Y^{n+1}_{k+1} &=& Y^{n+1}_k + \delta Y^{n+1}_{k+1}, \\
k &:=& k+1
\end{eqnarray*}
over $k$ until it converges. Then we set $Y^{n+1} := Y^{n+1}_{k}$. 
We use a combined stopping criterion 
that makes sure that both the norm of the residual vector $\Vert F(Y^{n+1})\Vert$ 
as well as the norm of the increment $\Vert \delta Y^{n+1}\Vert$ are sufficiently 
small.

@@

Automatic adaptivity in higher-order finite element methods (\emph{hp}-FEM) 
is much different from adaptivity in low-order FEM. Firstly, analytical
error estimates capable of guiding adaptive $hp$-FEM do not exist even for 
the simplest linear elliptic equations, not speaking about nonlinear multiphysics 
coupled systems. Secondly,
a higher-order element can be refined in many different ways, as illustrated 
in Fig.~\ref{fig:refinements}.
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.5\columnwidth]{refinements}
  \caption{\label{fig:refinements} Many possible refinement candidates for a fourth-order
  element.}
  \end{centering}
%\vspace{-0.5cm}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.8\columnwidth]{cand_list_quads}
  \caption{\label{fig:candlist} Refinement candidates for every
  refinement mode for quad type elements.}
  \end{centering}
\end{figure}
Note that triangular elements do not support anisotropic refinements.
Due to the large number of refinement options, classical error estimators 
that provide a constant error estimate per element, cannot be used to 
guide automatic \emph{hp}-adaptivity. 
For this, one needs to know the shape of the approximation error.
Hermes uses a pair of approximations with different orders of accuracy 
to obtain this information: coarse mesh solution and fine mesh solution~\cite{solin2010pde}. 
The initial coarse mesh is read from the mesh file, and the initial 
fine mesh is created through its global refinement both in \emph{h}
and \emph{p}. The fine mesh solution is the approximation of 
interest both during the adaptive process and at the end of computation. 
Global orthogonal projection of the fine mesh solution on the coarse mesh 
is used to extract the low-order part from the reference solution.
The adaptivity algorithm is guided by the difference between the 
reference solution and its low-order part.
Note that this approach to automatic adaptivity is PDE-independent
and thus naturally applicable to a large variety of multiphysics 
coupled problems.

>>>>

  X=\frac xl,\ \ \
  Y=\frac yl,\ \ \
  \tau=\frac{tD}{\lambda_D l},\ \ \

@@

  \frac{DC_0}{\lambda_D l}\frac{\partial c}{\partial \tau} + \frac 1L \nabla_d\cdot
    \left(-\frac{DC_0}{l}\nabla_dc-c\frac{DC_0}{l}\nabla_d\varphi\right) &=& 0\label{eq:dim1}\\
  - \frac{\varepsilon R T}{l^2F^2C_0}\nabla_d^2\varphi&=&c-1\label{eq:dim2},

@@

  \epsilon=\frac{\lambda_D}{l},

@@

  \frac{\partial c}{\partial \tau}
    -\epsilon\nabla_d^2 c - \epsilon \nabla_d \cdot \left( c\nabla_d \varphi\right)  &=&0,\label{eq:dimensionless-nernst-planck}\\
  -2\epsilon^2\nabla_d^2\varphi&=&c-1\label{eq:dimensionless-poisson}.

@@

$$H^1\left(\Omega\right)=\left\{v\in L^2\left(\Omega\right);\ \nabla_d v \in \left[L^2\left(\Omega\right)\right]^2\right\}.

@@

  \int_{\Omega}\frac{\partial c}{\partial \tau}v^c d\mathbf{x}-
  \int_{\Omega}\epsilon\nabla_d^2cv^c d\mathbf{x}-\int_{\Omega}\epsilon\nabla_d c\cdot\nabla_d\varphi v^c d\mathbf{x}-
  \int_{\Omega}\epsilon c\nabla_d^2\varphi v^c d\mathbf{x}=0.

@@

 \int_{\Omega}\frac{\partial c}{\partial \tau}v^c d\mathbf{x}+
  \epsilon\int_{\Omega}\nabla_d c\cdot\nabla_d v^c d\mathbf{x}-
  \epsilon\int_{\Omega}\nabla_d c\cdot\nabla_d\varphi v^c d\mathbf{x}

@@

  + \epsilon\int_{\Omega}\nabla_d\left(cv^c\right)\cdot \nabla_d \varphi d\mathbf{x}
  -\epsilon\int_{\partial\Omega}\frac{\partial c}{\partial n}v^c d\mathbf{S}-
  \int_{\partial\Omega}\epsilon\frac{\partial\varphi}{\partial n}cv^c d\mathbf{S}=0.

@@

\eqref{eq:dimensionless-nernst-planck-boundary}, we have

@@

  \int_{\Omega}\frac{\partial c}{\partial \tau}v^c d\mathbf{x}+
  \epsilon\int_{\Omega}\nabla_d c\cdot\nabla_d v^c d\mathbf{x}-
  \epsilon\int_{\Omega}\nabla_d c\cdot\nabla_d\varphi v^c d\mathbf{x}

@@

  + \epsilon\int_{\Omega}\nabla_d \varphi \cdot \nabla_d c v^c d\mathbf{x}+
  \epsilon \int_{\Omega} c \left(\nabla_d\varphi\cdot\nabla_d v^c\right) d\mathbf{x}=0.

@@

After the third and fourth terms cancel out, we obtain the final weak form of 

@@

  \int_{\Omega}\frac{\partial c}{\partial \tau}v^c d\mathbf{x}+
  \epsilon\int_{\Omega}\nabla_d c\cdot\nabla_d v^c d\mathbf{x}+
  \epsilon \int_{\Omega} c \left(\nabla_d\varphi\cdot\nabla_d v^c\right) d\mathbf{x}=0.

@@

Analogously we derive also the weak form of the Poisson equation (\ref{eq:dimensionless-poisson}),

@@

  -2\epsilon^2\int_{\Omega}\nabla_d^2\varphi v^\varphi d\mathbf{x}-\int_{\Omega}cv^\varphi d\mathbf{x}+
  \int_{\Omega}v^\varphi d\mathbf{x}=0.

@@

  -2\epsilon^2\int_{\Omega}\nabla_d\varphi\cdot\nabla_d v^\varphi d\mathbf{x}-\int_{\Omega}cv^\varphi d\mathbf{x}+
  \int_{\Omega}v^\varphi d\mathbf{x}=0.

@@

\subsection{Jacobian matrix and residual vector for the Newton's method}
To employ the Newton's method for the nonlinear system \eqref{eq:nernst-planck-weak-final},
\eqref{eq:poisson-weak-final}, formulas for the Jacobian matrix and residual vector need to
be derived. Time discretization will be performed using the second-order Crank-Nicolson 
method. The unknown solution components 
$c^{n+1}$ and $\varphi^{n+1}$ at the end of the time step $\delta\tau$ are expressed 
as linear combinations of finite element basis functions $v_k^{c}$ 
and $v_k^{\varphi}$ with unknown coefficients,
\begin{eqnarray}
  c^{n+1} = c(Y^{n+1}) &=& \sum_{k=1}^{N^c} y_k^{c} v_k^{c}, \label{eq:cnotation}\\
  \varphi^{n+1} = \varphi(Y^{n+1}) &=& \sum_{k=1}^{N^{\varphi}} y_k^{\varphi} v_k^{\varphi}\label{eq:phinotation}.
\end{eqnarray}
Here $Y^{n+1}$ is a coefficient vector of length $N^c + N^{\varphi}$ comprising the unknown 
solution coefficients $y_k^{C}$ and $y_k^{\varphi}$ (in this order). We will also be 
using $c^n = c(Y^n)$ and $\varphi^n = \varphi(Y^n)$ for the previous time step solutions.

With the notation~\eqref{eq:cnotation}, \eqref{eq:phinotation}, the time discretized 
Eq.~\eqref{eq:nernst-planck-weak-final} leads to the formula for the 
first part $F^c$ of the residual vector $F$,
\begin{eqnarray}
  F_i^c\left(Y\right) & = & \int_{\Omega} \frac{c(Y)}{\delta\tau}v_i^c d\mathbf{x} - 
  \int_{\Omega} \frac{c^{n}}{\delta\tau}v_i^c d\mathbf{x}\nonumber\\
  &&+\frac 12 \epsilon \left[\int_{\Omega} \nabla_d c(Y) \cdot \nabla_d v_i^c d\mathbf{x}+ 
  	\int_{\Omega} \nabla_d c^{n} \cdot \nabla_d v_i^c d\mathbf{x}\right.\nonumber\\
  &&+ \left.\int_{\Omega}c(Y) \left(\nabla_d \varphi(Y) \cdot \nabla_d v_i^c\right) d\mathbf{x}+
  \int_{\Omega}c^{n} \left(\nabla_d \varphi^{n} \cdot \nabla_d v_i^c\right) d\mathbf{x}\right]\label{eq:Fc}
\end{eqnarray}
where $i = 1, 2, \ldots, N^c$.
Analogously, Eq.~\eqref{eq:poisson-weak-final} defines the second part $F^{\varphi}$ 
of the residual vector $F$,
\begin{equation}
  F_i^{\varphi}\left(Y\right) = -2\epsilon^2\int_{\Omega} \nabla_d \varphi(Y) \cdot \nabla_d v_i^{\varphi} d\mathbf{x} 
  - \int_{\Omega} c(Y)v_i^{\varphi} d\mathbf{x} + \int_{\Omega} v_i^{\varphi} d\mathbf{x}
  \label{eq:Fphi}
\end{equation}
where $i = N^c + 1, N^c + 2, \ldots, N^c + N^{\varphi}$. The nonlinear discrete problem 
that needs to be solved at the end of each time step thus has the form 
$F(Y) = 0$.

%\newpage

\noindent
The Jacobian matrix $J(Y) = DF/DY$ has a $2\times 2$ block structure,
\begin{equation}
J(Y) = \left(
\begin{array}{cc}
  \displaystyle \frac{\partial F_i^c}{\partial y_j^c} & \displaystyle \frac{\partial F_i^c}{\partial y_j^{\varphi}} \\ 
  \displaystyle \frac{\partial F_i^{\varphi}}{\partial y_j^c} & \displaystyle \frac{\partial F_i^{\varphi}}{\partial y_j^{\varphi}}
\end{array}
\right),
\end{equation}
and its entries are obtained by calculating the partial derivatives of $F$ with
respect to the components of the coefficient vector $Y$. For this it is useful to 
realize that 
$$
\frac{\partial c(Y)}{\partial y_j^c} = v_j^c, \ \ \ 
\frac{\partial \nabla_d c(Y)}{\partial y_j^c} = \nabla_d v_j^c,\ \ \ \mbox{etc.}.
$$
We obtain
\begin{eqnarray}
  \frac{\partial F_i^c}{\partial y_j^c}(Y) &=& 
  \int_{\Omega} \frac{1}{\delta\tau} v_j^c v_i^c d\mathbf{x} + 
  \frac 12 \epsilon\left[\int_{\Omega} \nabla_d v_j^c \cdot \nabla_d v_i^c d\mathbf{x}
  + \int_{\Omega} v_j^c \left(\nabla_d \varphi(Y) \cdot \nabla_d v_i^c\right) d\mathbf{x}\right],\label{eq:dFcdyc}\\
  \frac{\partial F_i^c}{\partial y_j^{\varphi}}(Y) &=&
  \frac 12 \epsilon \int_{\Omega} c(Y) \left(\nabla_d v_j^{\varphi} \cdot \nabla_d v_i^c\right) d\mathbf{x},\\
  \frac{\partial F_i^{\varphi}}{\partial y_j^c}(Y) &=&
  - \int_{\Omega} v_j^c v_i^{\varphi} d\mathbf{x},\\
  \frac{\partial F_i^{\varphi}}{\partial y_j^{\varphi}}(Y) &=&
  -2\epsilon^2\int_{\Omega} \nabla_d v_j^{\varphi} \cdot \nabla_d v_i^{\varphi} d\mathbf{x}\label{eq:dFphidyphi}.
\end{eqnarray}

\subsection{Newton's iteration}
At the beginning of the $(n+1)$st time step we 
set $Y^{n+1}_0 = Y^n$, where $Y^n$ is the coefficient vector  
that was calculated in the $n$th time step (or coming from the 
initial condition if $n = 0$). We set $k = 0$ and run the 
Newton's iteration 
\begin{eqnarray*}
J(Y^{n+1}_k) \delta\tau Y^{n+1}_{k+1} &=& -F(Y^{n+1}_k), \\ 
Y^{n+1}_{k+1} &=& Y^{n+1}_k + \delta\tau Y^{n+1}_{k+1}, \\
k &:=& k+1
\end{eqnarray*}
over $k$ until it converges. Then we set $Y^{n+1} := Y^{n+1}_{k}$. 
We use a combined stopping criterion 
that makes sure that both the norm of the residual vector $\Vert F(Y^{n+1})\Vert$ 
as well as the norm of the increment $\Vert \delta Y^{n+1}\Vert$ are sufficiently 
small.

@@

% Pavel
% The reader might loose at this point the interest in reading further
% since there is no motivation which favros a hp-FEM discretization over traditional
% low order FEM. One sentence regarding computational performance
% or a related comment might be useful...
Automatic adaptivity in higher-order finite element methods (\emph{hp}-FEM) 
is much different from adaptivity in low-order FEM. Firstly, analytical
error estimates capable of guiding adaptive $hp$-FEM do not exist even for 
the simplest linear elliptic equations, not speaking about nonlinear multiphysics 
coupled systems. Secondly,
a higher-order element can be refined in many different ways, as illustrated 
in Fig.~\ref{fig:refinements}.
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.5\columnwidth]{refinements}
  \caption{\label{fig:refinements} Many possible refinement candidates for a fourth-order
  element.}
  \end{centering}
%\vspace{-0.5cm}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=0.8\columnwidth]{cand_list_quads}
  \caption{\label{fig:candlist} Refinement candidates for every
  refinement mode for quad type elements.}
  \end{centering}
\end{figure}
Note that triangular elements do not support anisotropic refinements.
Due to the large number of refinement options, classical error estimators 
that provide a constant error estimate per element, cannot be used to 
guide automatic \emph{hp}-adaptivity. 
For this, one needs to know the shape of the approximation error.
Hermes uses a pair of approximations with different orders of accuracy 
to obtain this information: coarse mesh solution and fine mesh solution~\cite{solin2010pde}. 
The initial coarse mesh is read from the mesh file, and the initial 
fine mesh is created through its global refinement both in \emph{h}
and \emph{p}. The fine mesh solution is the approximation of 
interest both during the adaptive process and at the end of computation. 
Global orthogonal projection of the fine mesh solution on the coarse mesh 
is used to extract the low-order part from the reference solution.
The adaptivity algorithm is guided by the difference between the 
reference solution and its low-order part.
% Pavel
% A short explanation, how Hermes further deals with this difference without
% going to much into details, might be interesting here...
Note that this approach to automatic adaptivity is PDE-independent
and thus naturally applicable to a large variety of multiphysics 
coupled problems.

### 2011-02-10 results.tex
Non-dimensionalized equations, some minor fixes

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step.

@@

\subsection{Comparison of single-mesh and multi-mesh $hp$-FEM}

Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted in
the smallest problems, shortest computing times, and better or the same error 
convergence compared to any single-mesh configuration.
This is illustrated for HP\_ANISO adaptivity mode in Figs.~\ref{fig:singlemultidof} 
and Fig.~\ref{fig:singlemulticpu}. The same holds true for HP\_ISO mode.

@@

\noindent 
Fig. \ref{fig:poly} shows higher-order meshes in the adaptive multimesh $hp$-FEM
computation for $C$ and $\phi$ at $t = 0.7$ s. Different 
colors mean different polynomial degrees. A diagonal pattern inside an element 
tells that the element has different polynomial degrees in the 
horizontal and vertical directions.

@@

These result are in good agreement with Fig.~\ref{fig:cphi-2} --- in the vicinity
of the boundaries $\partial \Omega_1$ and $\partial\Omega_3$, the concentration gradient
is much greater than the voltage gradient. Therefore the multimesh $hp$-FEM adaptivity 
algorithm has increased the maximum polynomial degree for the $C$-space to~7 while 
the maximum polynomial degree for the $\phi$-space is~2. One can also see that the mesh 
is significantly more refined for $C$.
Since these results are representative for all adaptivity modes, only multi-mesh 
configurations are considered in the following.

@@

To optimize the calculation time of HP\_ANISO, an adaptive time step control was employed.
The classical PID controller was used~\cite{valli2002control,dubcova2010space}.
Since $C$ and $\phi$ change differently in time as was demonstrated in Figs.~\ref{fig:cphi-1}
and~\ref{fig:cphi-2}, the relative changes between the solutions at different
time steps were monitored:
\begin{eqnarray}
  e_C^n & = & \frac{\lVert C^n-C^{n-1}\rVert}{\lVert C^n \rVert},\\
  e_\phi^n & = & \frac{\lVert \phi^n-\phi^{n-1}\rVert}{\lVert \phi^n \rVert}.
\end{eqnarray}
The relative changes to control the time step was calculated as follows:
\begin{equation}
  e^n=\max\left\{ e_C^n,\ e_\phi^n \right\}.
\end{equation}
If $e^n < \delta$ where $\delta>0$ is a defined tolerance, then the time
step for the next iteration is increased smoothly to
\begin{equation}
  \tau^{n+1}=\left( \frac{e^{n-1}}{e^n} \right)^{k_P}\ \left( \frac{\delta}{e^n} \right)^{k_l}\
  \left[ \frac{\left( e^{n-1} \right)^2}{e^n e^{n-2}} \right]^{k_D} \tau^n,
\end{equation}
where parameters are from~\cite{valli2002control}:
\begin{equation}
  k_p=0.075,\quad k_l=0.175,\quad k_D=0.01.
\end{equation}
The tolerance $\delta$ was set to $\delta=0.25$ in the current optimization
example. The calculated $e_C^n$ and $e_\phi^n$ are shown in Fig.~\ref{fig:cphirelerr}.
The HP\_ANISO problem size and computing time with and without time step control
are shown in Figs.~\ref{fig:timeadapt_dof} and~\ref{fig:timeadapt_cpu}. The reader
can notice that the computing time was reduced approximately 3 times when the time
step control was employed.

>>>>

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step. The dimensional variables $x$, $y$, $t$, $C$,
and $\phi$ are used to present the solutions.

@@

\subsection{Comparison of single-mesh and multi-mesh $hp$-FEM}
% David: Error convergence plots here.
Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted in
the smallest problems, shortest computing times, and better or the same error 
convergence compared to any single-mesh configuration.
This is illustrated for HP\_ANISO adaptivity mode in Fig.~\ref{fig:singlemultidof}, Fig.~\ref{fig:singlemulticpu} and Fig.~\ref{fig:singlemultierror}.
The same holds true for HP\_ISO mode.

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_error}
  \caption{\label{fig:singlemultierror} Relative solution error as a function 
  of physical time for single-mesh and multi-mesh configurations with 
  HP\_ANISO adaptivity mode.}
  \end{centering}
\end{figure}
\noindent 
Fig. \ref{fig:poly} shows higher-order meshes in the adaptive multi-mesh $hp$-FEM
computation for $C$ and $\phi$ at $t = 0.7$ s. Different 
colors mean different polynomial degrees. A diagonal pattern inside an element 
tells that the element has different polynomial degrees in the 
horizontal and vertical directions.

@@

The result are in good agreement with Fig.~\ref{fig:cphi-2} --- in the vicinity
of the boundaries $\partial \Omega_1$ and $\partial\Omega_3$, the concentration gradient
is much greater than the voltage gradient. Therefore the multi-mesh $hp$-FEM adaptivity 
algorithm has increased the maximum polynomial degree for the $C$-space to~7 while 
the maximum polynomial degree for the $\phi$-space is~2. One can also see that the mesh 
is significantly more refined for $C$.
Since these results are representative for all adaptivity modes, only multi-mesh 
configurations are considered in the following.

@@

To optimize the calculation time of HP\_ANISO, an adaptive time step control was employed.
The classical PID controller was used~\cite{valli2002control,dubcova2010space}.
Since $C$ and $\phi$ change differently in time as was demonstrated in Figs.~\ref{fig:cphi-1}
and~\ref{fig:cphi-2}, the relative changes between the solutions at different
time steps were monitored:
\begin{eqnarray}
  e_C^n & = & \frac{\lVert C^n-C^{n-1}\rVert}{\lVert C^n \rVert},\\
  e_\phi^n & = & \frac{\lVert \phi^n-\phi^{n-1}\rVert}{\lVert \phi^n \rVert}.
\end{eqnarray}
The relative changes to control the time step was calculated as follows:
\begin{equation}
  e^n=\max\left\{ e_C^n,\ e_\phi^n \right\}.
\end{equation}
If $e^n < \delta$ where $\delta>0$ is a defined tolerance, then the time
step for the next iteration is increased smoothly to
\begin{equation}
  \delta\tau^{n+1}=\left( \frac{e^{n-1}}{e^n} \right)^{k_P}\ \left( \frac{\delta}{e^n} \right)^{k_l}\
  \left[ \frac{\left( e^{n-1} \right)^2}{e^n e^{n-2}} \right]^{k_D} \delta\tau^n,
\end{equation}
where parameters are from~\cite{valli2002control}:
\begin{equation}
  k_p=0.075,\quad k_l=0.175,\quad k_D=0.01.
\end{equation}
The tolerance $\delta$ was set to $\delta=0.25$ in the current optimization
example. At this point, the implementation does not support
reducing the time step when $e^n\geq \delta$. The calculated $e_C^n$ and $e_\phi^n$ are shown in Fig.~\ref{fig:cphirelerr}.
The HP\_ANISO problem size and computing time with and without time step control
are shown in Figs.~\ref{fig:timeadapt_dof} and~\ref{fig:timeadapt_cpu}. The reader
can notice that the computing time was reduced approximately 3 times when the time
step control was employed.
